HO unification from object language to meta language

### 2024-02-22 paper/main.tex
update tex

In order to reason about unification of the terms of the objet language within the
meta language, we start by formally describing the two languages. Employing
meta-programming for this purpose, \cref{code:lang-descr} presents the type
\elpiIn{tm} containing the constructors for term application, lambda abstraction
and constants. Moreover, in order to represent unification variables, we need 
to give two different constructors.

\

In the case of the OL, variables have no scope. For example, the subterm
\coqIn{P x} from the instance forall\_dec. Here, \coqIn{P} is an higher order
variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} is a name bound to
\coqIn{P}. However, at the meta level, the translation of \coqIn{P x} becomes
\elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
encoding, unification variables are encoded as integer corresponding to memory
addresses, and the constructor for unification variables in the OL is
\elpiIn{fo_uv}. This constructor is prefixed with \elpiIn{fo} since the
variable, having no scope, is a first-order variable.

On the other hand, in the case of the ML, we want to use its unification
algorithm to make variable assignment. Since the ML is an Higher Order
Programming Language, we represent unification variables with the \elpiIn{uv}
constructor, which, this time, can see a list of terms. Of course some attention
should be payed when dealing with this constructor, since we have to certify
each time that an \elpiIn{uv i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is a list of distinct names. Finally,
the following code

@@

\noindent
illustrates the \elpiIn{assmt} representing variable assignment in the ML. \marginpar{todo: explain better}

The memory of the two languages are represented with lists of substitutions. In
particular, the substitution of the OL, called \elpiIn{fo_subst}, is made by
optional terms such that, if the substitution is \elpiIn{none}, then the
variable is not instantiated. Note that the variables in the \elpiIn{fo_subst}
have always the \elpiIn{fo_uv} constructor. On the other hand, the ML
substitution is an optional assignment and in that assignment, variables are
considered to have the \elpiIn{uv} constructor.

>>>>

In order to reason about unification of the terms of the objet language within the
meta language, we start by formally describing the two languages. Employing
meta-programming for this purpose, we present in \cref{code:common-terms} a new type for the terms of
the OL.
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{elpicode}
    kind tm type.
    type app list tm -> tm.
    type lam (tm -> tm) -> tm.
    type c   string -> tm.
  \end{elpicode}
  \caption{Common terms}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
%
\noindent
This encoding is very similar to the one introduced in \cref{sec:intro},\todo{say we do not care about types} except
for the \elpiIn{all} constructor. We explicitly neglect it since blabla.
Moreover, since we are working with unification variables we need to introduce a
new constructor.

In the case of the OL, variables have no scope. For example, the subterm
\coqIn{P x} from the instance forall\_dec has \coqIn{P} an higher order
variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} as bound variable. 
However, at the meta level, the translation of \coqIn{P x} becomes
\elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
encoding, we represent the variables of the OL in the following way:

\begin{elpicode}
  type fo_uv nat -> tm.
\end{elpicode}

\noindent 
In particular, a variable of the OL is identified by a integer
refereeing a memory address. This constructor is prefixed with \elpiIn{fo} since
the variable, having no scope, is a first-order variable. Finally, the memory of
the OL is depicted by the following type abbreviation:

\begin{elpicode}
  typeabbrev fo_subst list (option tm).
\end{elpicode}

\noindent 
which is a list of optional terms. If a cell is \elpiIn{none}, then the variable
corresponding to this cell is not instantiated.

On the other hand, the terms of the ML have exactly the same shape
has the terms of \cref{code:common-terms}. However, since the ML is an Higher
Order Programming Language, unification unification variables must have a scope.
We represent them in the following way:

\begin{elpicode}
  type ho_uv nat -> list tm -> tm.
\end{elpicode}

Of course some attention
should be payed when dealing with \elpiIn{ho_uv}, since we have to certify
that an \elpiIn{uv i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is always a list of distinct names.

Another important implementation should be provided in order to represent the
lambda abstractions and the applications of the ML. As outlined in
\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not the same semantic as 
<<\elpiIn{lam x\ app[f, x]}>>, since, even though, the two terms represent the same
concept, the former is the lambda abstraction of \elpiIn{f} applied to the 
binder in ML, whereas the latter is the same term representation but wrt the OL.
We already have defined the nodes \elpiIn{app} and \elpiIn{lam} for the OL,
therefore, in the following code snippet, we give their representation in the
ML.

@@

\noindent
In particular, the node \elpiIn{abs} stands for the lambda abstraction. The
node \elpiIn{val} contains terms of the object language. The node for variable
application is not really necessary, since, using the syntax of the ML, it is
defined by putting two terms of the ML side by side. For example, the terms 
<<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.

The memory of the ML, is defined with the following type abbreviation:

\begin{elpicode}
  typeabbrev ho_subst list (option tm).
\end{elpicode}

\noindent
with the invariant that the \elpiIn{tm} inside the cell never contains the node
\elpiIn{fo_uv}. Note that, reciprocally, the \elpiIn{fo_subst} does not contain
any node of the form \elpiIn{ho_uv}.

\subsection{Term equality}

\old
